A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-22 paint-contour-tree.tex

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

Consider the set of manifolds output by $\rain(x,\MM)$, there is a natural tree ordering defined over these manifolds where the 
root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).

%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
%the contour appears as a boundary maxima/minima.

\begin{claim} \label{clm:rain-reeb} Let $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
Then $\reeb(\MM)$ can be computed in $O(n)$ time.
\end{claim}

\begin{proof}
We will walk through the Manifold Ordering Tree in a leaf first ordering.  Each time we visit a node we connect its reeb graph to the 
reeb graph of its children in the tree using the $\surgery$ procedure. This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).  The correctness follows by induction and  \Lem{surgery}.
\end{proof}

\section{Painting to compute contour trees} \label{sec:paint}
The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is a \emph{minimum}.
We will fix such a manifold $\MM$, with the dominant minimum $u$.

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
%the contour appears as a boundary maxima/minima. 
}

\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
\end{claim}

\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
labeled with some $\MM_i$.
%
%There is a natural tree ordering defined over these manifolds where the 
%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
contour tree of its children in the tree using the $\surgery$ procedure. 
Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
%
%
%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).  The correctness follows by induction and  \Lem{surgery}.
\end{proof}

\section{Contour trees of extremum dominant manifolds} \label{sec:paint}

The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
We first explain what these trees are.

For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
Conventially, all edges are directed from higher to lower function value.

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
\end{definition}

Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.

The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

\begin{lemma} \label{lem:split} $\cS(\MM)$ consists of:
\begin{asparaitem}
	\item A single path (in sorted order) with all vertices except non-dominant minima.
	\item Each non-dominant minimum is attached to a unique split.
\end{asparaitem}
\end{lemma}

\begin{proof} First, let us complete the proof assuming that each split $v$ has one child that is just a leaf, which
is a non-dominant minimum. Any minimum is a leaf in $\cS(\MM)$, and thereby attached to a split. If we removed
all non-dominant minima, we must end up with a path, as asserted above.

Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
on this path. It must intersect some contour, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
contains $m^*$ as the non-ascending path $\phi$ intersects $\phi$ only once. Let the other
component be called $\MM'$.

Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
minimum.

The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
be the non-dominant minimum $w$, which is a leaf. 
\end{proof}

It is convenient to denote the non-dominant minima by $m_1, m_2, \ldots, m_k$, and the associated
splits (as given by \Lem{split}) by $s_1, s_2, \ldots, s_k$.

To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
from the join and split trees. 
It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
is the contour tree with regular points inserted into the edges. 
For any tree $T$ with vertex $v$, we define \emph{erasing} $v$ from $T$,
when both in and outdegrees are at most $1$. If $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\merge(\cJ(\MM),\cS(\MM))$}

\smallskip
\begin{compactenum}
	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
	\item Add all candidates to queue.
	\item While candidate queue is non-empty: 
	\begin{compactenum}
		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
		consider edge in $\cS$. In either case, let edge by $(v,w)$.
		\item Insert $(v,w)$ in $\cA(\MM)$. 
		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
	\end{compactenum}
	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
\end{compactenum}
\end{minipage}}

\medskip

\begin{theorem} \label{thm:contour-tree} The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$
as well as: for any critical point $v$, edge $(w,v)$ where $w$ is the lowest non-minimal critical point
in a connected component of $\MM^+_v$ connected to $v$.
\end{theorem}

\begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
maxima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
The join tree is now $\cJ(\MM) \ominus \{m_i\}$. We can now process $\cJ$ leaf by leaf, and all
edges of $\cJ$ are inserted into $\cA(\MM)$.

Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Smoothing out regular
points from $\cJ(\MM) \ominus \{m_i\}$ exactly yields the edges described in the theorem.
\end{proof}

\section{Painting to compute contour trees}
